Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-22 doc/asplos13/intro.tex
fixed up intro

Irregular applications generates tasks with work, interdependences, or memory
accesses that are highly sensitive to input. Classic examples of irregular
applications include branch and bound optimization, SPICE circuit simulation,
contact algorithms in car crash analysis, and network flow. Important
contemporary examples include processing large graphs in the business,
national security, machine learning, data-driven science, and social network
computing domains. For these emerging applications, fast response -- given the
sheer amount of data -- requires multinode systems. The most broadly available
multinode systems are those built from x86 commodity computing nodes
interconnected via ethernet or infiniband. Our goal is to enable scalable
performance of irregular applications on these mass market systems.

@@

\noindent{\bf Irregular applications observe little spatial locality.} It is
not atypical for any given task's data references to be spread randomly across
the entire memory of the system. This makes current memory hierarchy features
ineffective. Unfortunately, commodity prefetching hardware is effective only
when addresses are known many cycles before the data is consumed or the
accesses follow a predictable pattern, neither of which occurs in irregular
applications. As a consequence, commodity microprocessors stall often when
executing irregular applications.

% What makes our goal challenging is that irregular applications do not
% readily exploit the features of commodity systems.	This is true even for
% single node systems.	It is not atypical for any given task's data
% references to be spread randomly across the entire memory of the system.
% Yet, for commodity processors' caches to be effective, computations must
% exploit spatial locality.	Unfortunately, commodity prefetching hardware is
% effective only when addresses are known many cycles before the data is
% consumed or the accesses follow a predictable pattern, neither of which
% occurs in irregular applications.	As a consequence, commodity
% microprocessors stall often when executing irregular applications.

\noindent{\bf Irregular applications frequently request off-node data with
small messages.} On multinode systems, the challenges presented by low
locality are analogous, and exacerbated by the increased latency of going
off-node. Irregular applications also present a challenge to mass market
network technology, which is designed to transfer large blocks of data, not
the word-sized references emitted by irregular application tasks. Injection
rate into the network is insufficient to utilize wire bandwidth when blocks
are below about two-kilobytes, so any straightforward communication strategy
severely under-utilizes the network.

@@

Luckily, many of the important irregular applications (e.g., graph processing,
our focus on this paper) naturally offer large amounts of concurrency. This
immediately suggests taking advantage of concurrency tolerate latencies of
dealing with data movement. The fully custom Tera MTA-2~\cite{mta-2} system is
classic example of supporting irregular applications by using concurrency to
hide latencies. The MTA was well-suited to extremely irregular applications:
it had a large distributed shared memory with no caches, relying on
multithreaded processors to tolerate memory latency. On every clock cycle,
each processor would execute a ready instruction chosen from one of its 128
hardware thread contexts. Any application with sufficient thread parallelism
would fully utilize the processors independent of the memory access pattern
(modulo hot-spots). The network was designed with an single word injection
rate that matched the processor clock frequency and sufficient bandwidth to
sustain a reference from every processor on every clock cycle. Unfortunately,
while an excellent match to extremely irregular applications, the MTA was not
cost-effective on applications that could exploit locality and had very poor
single threaded performance, making it a commercial failure. The Cray XMT
approximates the Tera MTA-2, substantially reducing its cost but not
overcoming its narrow range of applicability.

@@

In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.  This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating with inexpensive mass market hardware components the approach taken by Tera. We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.  Ultimately, the opportunity is to cover the spectrum of irregular to regular computation:  where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.  Thus we make best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.

In this paper we introduce Grappa, a software runtime system that allows a
commodity x86 distributed-memory HPC cluster to be programmed as if it were a
single large shared-memory machine and provides scalable performance for
irregular applications. Grappa is designed to smooth over some of the
performance discontinuities in commodity hardware, giving good performance
when there is little locality to be exploited while allowing the programmer to
exploit it when it is available. Grappa leverages as much freely available and
commodity infrastructure as possible. We use unmodified Linux for the
operating system. An off the shelf user-mode infiniband device driver
stack~\cite{Melonox?} is used for network interfacing. MPI~\cite{mpi} is used
for process setup and tear down. GASnet~\cite{gasnet} is used as the
underlying mechanism for remote memory reads and writes using active message
invocations. To this commodity hardware and software mix Grappa adds three
main software components: (1) a \emph{lightweight tasking} layer that supports a
context switch in as few as \checkme{40ns} and distributed load balancing; (2)
a \emph{distributed shared memory} layer that supports customary operations such as
\emph{read and write as well as remote operations} such as
fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message aggregation} layer that
combines short messages to mitigate the aforementioned problem that commodity
networks are designed to achieve peak bandwidth only on large packet sizes,
yet irregular applications tend to fetch only a handful of bytes at a time.

>>>>

Irregular applications generate tasks with work, interdependences, or memory
accesses that are highly sensitive to input. Classic examples of irregular
applications include branch and bound optimization, SPICE circuit simulation,
contact algorithms in car crash analysis, and network flow. Important
contemporary examples include processing large graphs in the business,
national security, machine learning, data-driven science, and social network
computing domains. For these emerging applications, fast response -- given the
sheer amount of data -- requires multinode systems. The most broadly available
multinode systems are those built from x86 commodity computing nodes
interconnected via ethernet or infiniband. Our goal is to enable scalable
performance of irregular applications on these mass market systems.

@@

\noindent{\bf Irregular applications exhibit little spatial locality.}
It is not atypical for any given task's data references to be spread
randomly across the entire memory of the system. This makes current
memory hierarchy features ineffective. Caches are of little assistance
with such low data re-use and spatial locality. Commodity prefetching
hardware is effective only when addresses are known many cycles before
the data is consumed or the accesses follow a predictable pattern,
neither of which occurs in irregular applications. As a consequence,
commodity microprocessors stall often when executing irregular
applications.

\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low
locality are analogous, and exacerbated by the increased latency of going
off-node. Irregular applications also present a challenge to mass market
network technology, which is designed to transfer large blocks of data, not
the word-sized references emitted by irregular application tasks. Injection
rate into the network is insufficient to utilize wire bandwidth when blocks
are below about two-kilobytes, so any straightforward communication strategy
severely under-utilizes the network.

@@

Luckily, many of the important irregular applications (e.g., graph
processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency
to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{mta-2} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches.  On every clock cycle, each
processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory
access latency.  The network was designed with a single word injection
rate that matched the processor clock frequency and sufficient bandwidth
to sustain a reference from every processor on every clock cycle.
Unfortunately, while an excellent match to extremely irregular
applications, the MTA was not cost-effective on applications that could
exploit locality and had very poor single threaded performance, making
it a commercial failure. The Cray XMT approximates the Tera MTA-2,
substantially reducing its cost but not overcoming its narrow range of
applicability.

@@

In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well.  This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera. We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth.  Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency.  Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.

In this paper we introduce Grappa, a software runtime system that allows
a commodity x86 distributed-memory HPC cluster to be programmed as if it
were a single large shared-memory machine and provides scalable
performance for irregular applications. Grappa is designed to smooth
over some of the performance discontinuities in commodity hardware,
giving good performance when there is little locality to be exploited
while allowing the programmer to exploit it when it is available. Grappa
leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system. An off the
shelf user-mode infiniband device driver stack~\cite{Melonox?} is used
for network interfacing. MPI~\cite{mpi} is used for process setup and
tear down. GASnet~\cite{gasnet} is used as the underlying mechanism for
remote memory reads and writes using active message invocations. To this
commodity hardware and software mix Grappa adds three main software
components: (1) a \emph{lightweight tasking} layer that supports a
context switch in as few as 38ns and distributed load balancing; (2) a
\emph{distributed shared memory} layer that supports customary
operations such as \emph{read and write as well as remote operations}
such as fetch-and-add~\cite{fetchandadd}; and (3) a \emph{message
aggregation} layer that combines short messages to mitigate the
aforementioned problem that commodity networks are designed to achieve
peak bandwidth only on large packet sizes, yet irregular applications
tend to fetch only a handful of bytes at a time.
